Local Spectral Clustering of Density Upper Level Sets

### 2019-05-14 arxiv/arxiv_local_density_clustering.tex
updated proofs, detailed edits of arxiv version. Copied/pasted neurips version

\documentclass{article}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{float}
\usepackage[export]{adjustbox}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage[reqno]{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[parfill]{parskip}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{bm}
\usepackage{fullpage}

@@

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
\newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\mathbf{\Phi}}
\newcommand{\Psibf}{\mathbf{\Psi}}
\newcommand{\dist}{\mathrm{dist}}

%%% Vectors
\newcommand{\pbf}{\mathbf{p}}
\newcommand{\qbf}{\mathbf{q}}
\newcommand{\ebf}[1]{\mathbf{e}_{#1}}
\newcommand{\pibf}{\bm{\pi}}

%%% Matrices
\newcommand{\Abf}{\mathbf{A}}
\newcommand{\Xbf}{\mathbf{X}}
\newcommand{\Wbf}{\mathbf{W}}
\newcommand{\Lbf}{\mathbf{L}}
\newcommand{\Dbf}{\mathbf{D}}
\newcommand{\Ibf}[1]{\mathbf{I}_{#1}}

@@

%%% Sets
\newcommand{\Cset}{\mathcal{C}}
\newcommand{\Aset}{\mathcal{A}}
\newcommand{\Asig}{\Aset_{\sigma}}
\newcommand{\Csig}{\Cset_{\sigma}}

@@

\begin{abstract}
\vskip 0.1 in % RESOLVED
Spectral clustering methods are a family of popular nonparametric clustering
tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
which extract clusters using locally-biased random walks around a user-specified
seed node.  Several authors have shown that local methods, such as personalized
PageRank (PPR), have worst-case guarantees for certain graph-based measures of
cluster quality.  In contrast to existing works, we analyze PPR in a traditional
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We introduce two natural criteria for cluster quality, and derive
bounds for these criteria when evaluated on empirical analogues of density
clusters. Moreover, we prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters.
\end{abstract}

@@

Let $\mathbf{X} = \{x_1, \ldots, x_n\}$ be a sample drawn i.i.d.\ from a
distribution $\Pbb$ on $\Rd$, with density $f$, and consider the problem of 
clustering: splitting the data into groups which satisfy some notion of
within-group similarity and between-group difference.  We focus on spectral
clustering methods, a family of powerful nonparametric clustering algorithms.
Roughly speaking, a spectral technique first constructs a geometric graph $G$,
where vertices are associated with samples, and edges correspond to proximities
between samples. It then learns a feature embedding based on the Laplacian of
$G$, and applies a simple clustering technique (such as k-means clustering) in
the embedded feature space.

To be more precise, let $G=(V,E,w)$ denote a weighted, undirected graph  
constructed from the samples $\mathbf{X}$, where $V=\{1,\ldots,n\}$, and $w_{uv}
= K(x_u,x_v) \geq 0$ for $u,v \in V$, and a particular kernel function $K$.
Here $(u,v) \in E$ if and only if $w_{uv} > 0$.  We denote by $\Abf \in
\Reals^{n \times n}$ the weighted adjacency matrix, which has entries
$A_{uv}=w_{uv}$, and by $\Dbf$ the degree matrix, with 
\smash{$\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$}.  We also denote by $\Wbf,\Lbf$
the random walk transition probability matrix and normalized\footnote{Other
  popular choices here include the unnormalized Laplacian, and symmetric
  normalized Laplacian.} 
Laplacian matrix, respectively, which are defined as
$$
\Wbf = \Dbf^{-1}\Abf, \quad \Lbf = \Ibf{} - \Wbf,
$$
where $\Ibf{} \in \Reals^{n\times n}$ is the identity matrix.  Classical global
spectral methods take a eigendecomposition $L=U \Sigma U^T$, use some 
number of eigenvectors (columns in $U$) as a feature representation for the
samples, and then run (say) k-means in this new feature space.

@@

Define the cut between subsets $S,S' \subseteq V$ to be
$$
\cut(S, S'; G) = \sum_{u \in S} \sum_{v \in S'} w_{uv},
$$
and define
\smash{$\vol(S; G) = \cut(S, V; G) = \sum_{u \in S} \sum_{v \in V} w_{uv}$}.  
As our notion of external connectivity, we use the \emph{normalized cut}  
of $S$, defined as 
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) = \frac{\cut(S; G)}{\min\{\vol(S; G), \vol(S^c; G)\}},
\end{equation} 
where we abbreviate $\cut(S; G) = \cut(S; S^c; G)$.

@@

For $S \subseteq V$, denote by $G[S] = (S, E_S, w_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Abf_S,\Dbf_S$
be the adjacency matrix and degree matrix, respectively, of $G[S]$.  Define the
random walk matrix as usual, $\Wbf = \Dbf_S^{-1} \Abf_S$, and for $v \in V$,
write 
$$
q_{vu}^{(t)} = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of a random walk over $G[S]$
originating at $v$.\footnote{Given a starting node $v$ and and a random walk
  defined by transition probability matrix $\mathbf{P}$, the notation $e_v
  \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$
  steps.}   Also write \smash{$\widetilde\pibf = (\widetilde\pi_{u})_{u \in S}$}
for the stationary distribution of this random walk.  (Given the definition of 
$\Wbf_S$, it is well-known that the stationary distribution is given by
\smash{$\widetilde\pi_{u} = (\Dbf_{S})_{uu} / \vol(S; G[S])$}.)

Our internal connectivity parameter will capture the time it takes for the
random walk over $G[S]$ to mix (approach the stationary distribution)
uniformly over $S$.  For this, we first define the \emph{relative pointwise mixing
  time} of $G[S]$ as 
$$
\tau_{\infty}(G[S]) = \min\set{ m: \frac{|q_{vu}^{(m)} - 
      \widetilde\pi_u|}{\widetilde\pi_u} \leq \frac{1}{4}, 
    \; \text{for $u,v \in V$}}. 
$$
%where \smash{$\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}$}, and
%\smash{$\qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$}. 
% RJT: I got rid of the notational dependence on \qbf, because these quantities
% are in turn just defined in terms of the induced subgraph G[S]
% AJG: RESOLVED
Now our internal connectivity parameter is simply the inverse mixing time, 
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(G[S])}.
\end{equation}

@@

We now describe the clustering algorithm that will be our focus for the rest of 
the paper. We start with the geometric graph that we form based on the samples 
$\Xbf$: for a radius $r > 0$, we consider the \emph{$r$-neighborhood graph} of 
$\Xbf$, denoted $G_{n,r}=(V,E)$, an unweighted graph with vertices
$V=\{1,\ldots,n\}$, and an edge $(u,v) \in E$ if and only if $\norm{x_u - x_v}
\leq  r$, where $\norm{\cdot}$ denotes Euclidean norm.  Note that this is a
special case of the general construction introduced above, with 
$K(u,v) = 1(\norm{x_u - x_v} \leq r)$.

@@

For a level $\beta > 0$ and a target volume $\vol_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: \textcolor{red}{\frac{p_u}{\Dbf_{uu}}} > \frac{\beta}{\vol_{0}}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}
we output a cluster \smash{$\widehat{C} = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a Neighborhood Graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation 
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target stationary volume $\pi_0 >
0$. \\   
{\bfseries Output:} cluster $\widehat{C} \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn:
    ppr_vector}. 
  \STATE For $\beta \in (\frac{1}{40}, \frac{1}{11})$ compute sweep cuts
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return \smash{$\widehat{C} = S_{\beta^*}$}, where 
  $$
  \beta^* = \argmin_{\beta \in (\frac{1}{40}, \frac{1}{11})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

@@

\begin{definition}[Misclassification rate]
\label{def: misclassification_rate}
For an estimator \smash{$\widehat{\Cset} \subseteq \Xbf$} and set
$\mathcal{S} \subseteq \Reals^d$, the \emph{misclassification rate} of $\mathcal{S}$ by $\widehat{\Cset}$ is
\begin{equation}
\label{eqn: misclassification_rate}
\abs{\widehat{\Cset} \setminus (\mathcal{S} \cap \Xbf)} + \abs{(\mathcal{S} \cap \Xbf) \setminus \widehat{\Cset}}.
\end{equation}
\end{definition}

\begin{definition}[Consistent density cluster estimation]
\label{def: consistent_density_cluster_estimation}
For an estimator \smash{$\widehat{\Cset} \subseteq \Xbf$} and cluster 
$\Cset \in \Cbb_f(\lambda)$, we say \smash{$\widehat{\Cset}$} is a consistent
estimator of $\Cset$ if for all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset \not=
\Cset'$ the following holds as $n \to \infty$: 
\begin{equation}
\label{eqn: consistent_density_cluster_recovery}
\Cset[\Xbf] \subseteq \widehat{\Cset} \quad \text{and} \quad
\widehat{\Cset} \cap \Cset'[\Xbf] = \emptyset,
\end{equation}
with probability tending to 1.
\end{definition}

@@

\item Under \textcolor{red}{largely the same set of geometric conditions}, we derive in Theorem
  \ref{thm: inverse_lower_bound_nonconvex} a lower bound on the
  inverse mixing time of a random walk over $\Csig[\Xbf]$.

\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we show in
  Theorem \ref{thm: consistent_recovery_of_density_clusters} that the bounds on the cluster quality criteria established in Theorems \ref{thm: conductance_upper_bound} and \ref{thm:
    inverse_mixing_time_lower_bound_nonconvex} have algorithmic consequences for \ppr: 
  properly initialized, Algorithm \ref{alg: ppr} has low misclassification rate with respect to a small enlargement of the set $\Cset$, and if the density cluster $\Cset$ is particularly well-conditioned, Algorithm \ref{alg: ppr} will perform consistent density cluster estimation in the sense of \eqref{eqn: consistent_density_cluster_recovery}.

\item We show in Corollary \ref{cor: appr} that these statements hold also with respect to an approximate form of \pprspace, which can be efficiently computed.

\item In Section \ref{sec: experiments}, we empirically
  demonstrate the tightness of the bounds in Theorems \ref{thm: conductance_upper_bound} and \ref{thm: inverse_mixing_time_lower_bound}, and provide examples showing how violations of the geometric conditions we require manifestly
  impact density cluster recovery by \ppr.  
\end{enumerate}

@@

In addition to the background given above, a few related lines of work are worth
highlighting.
Global spectral clustering methods were first developed in the
context of graph partitioning \citep{fiedler1973,donath1973} and their
performance is well-understood in this context (see, e.g.,
\citealt{tolliver2006,luxburg2007}).  In a similar vein, several recent works
\citep{mcsherry2001,rohe2011,kamalika2012,balakrishnan2011,lei2015,abbe2018} 
have studied the efficacy of spectral methods in successfully recovering the
community structure in the stochastic block model and variants.

@@

\item
\label{asmp: embedding}
\emph{Lipschitz embedding:}
There exists $\mathcal{K} \subseteq \Rd$ convex, and $g: \Reals^d \to \Reals^d$ satisfying
\begin{equation*}
\det(\nabla g (x) ) = 1,\frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~ \text{for all $x,y \in \Reals^d$}
\end{equation*}
such that $\Csig$ is the image of $\mathcal{K}$ by $g$, $\Csig = g(\mathcal{K})$.
Furthermore, there exists $D < \infty$ such that for all $x, x' \in \mathcal{K}$
$$
\norm{x - x'} \leq D.
$$
\end{enumerate}

Note that $\sigma$ plays several roles here, precluding arbitrarily narrow
clusters and long clusters in \ref{asmp: bounded_density}, flat densities around the level set in \ref{asmp: low_noise_density}, and poorly separated clusters in \ref{asmp: cluster_separation}.

Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density},
and \ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf];
G_{n,r})$, whereas \ref{asmp: bounded_density}, and \ref{asmp: embedding} are required to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the
lower bound on minimum density in \ref{asmp: bounded_density} along with \ref{asmp:  
cluster_separation} are similar to the $(\sigma,\epsilon)$-saliency of
\citep{chaudhuri2010}, a standard density clustering assumption, while
\ref{asmp: low_noise_density} is seen in, e.g., \citep{singh2009} (as well as
many other works on density clustering and level set estimation.)  It is worth
highlighting that these assumptions are all local in nature, a benefit of
studying a local algorithm such as \ppr.

@@

We also note that while many of these geometric conditions are typical in the density clustering literature, the restrictions we will impose upon them in order to obtain meaningful implications for \pprspace will not be. This is natural. The spectral algorithm we consider is not specifically designed for the task of level set estimation, and in fact one should expect \pprspace to fail to recover -- either in the sense of \eqref{eqn: consistent_density_cluster_recovery}, or indeed any reasonable notion of cluster recovery -- a density cluster of sufficiently large diameter or sufficiently small thickness (though we do not provide any lower bounds to this effect). Indeed, one of the primary motivations of this work was to better understand and characterize the distinctions between those level sets which are well conditioned for spectral algorithms, and those which are not.

@@

\begin{remark}
The inequality in \eqref{eqn: conductance_additive_error_bound} is tight in the case of $\Cset = \set{0}$. To see this, let $\Csig = B(0,\sigma)$ and
$$
 f(x)  = \begin{cases}
   \lambda &\text{for $x \in \Csig$}, \\
   \lambda - \dist(x,\Csig)^{\gamma} & \text{ for $0 < \dist(x,\Csig) < r$}, 
\end{cases}
$$
Then, some simple calculations yield
\begin{equation*}
\frac{\Pbb((\Csig + B(0,r)) \setminus \Csig)}{\Pbb(\Csig)} \geq d\frac{(\lambda - r^{\gamma}) r}{\sigma},
\end{equation*}
and for some $c > 0$,
\begin{equation*}
\Ebb(\cut(\Csig[\Xbf]; G_{n,r})) \setminus \Csig) \geq c \lambda \nu_d r^d \Pbb((\Csig + B(0,r)), \quad \textrm{and} \quad \Ebb(\vol_{n,r}(\Csig[\Xbf]; G_{n,r}) \leq c \lambda \nu_d r^d \Pbb(\Csig)
\end{equation*}
so the ratio $\Ebb(\cut(\Csig[\Xbf]; G_{n,r})) / \Ebb(\vol_{n,r}(\Csig[\Xbf]; G_{n,r})$ matches \ref{eqn: conductance_additive_error_bound}, up to constants.

>>>>

\documentclass{article}
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{float}
\usepackage[export]{adjustbox}
\usepackage{subcaption}
\usepackage{booktabs}
\usepackage{xcolor}
\usepackage{xr-hyper}
\usepackage{hyperref}
\usepackage[reqno]{amsmath}
\usepackage{amsfonts, amsthm, amssymb}
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage[parfill]{parskip}
\usepackage{enumerate}
\usepackage[shortlabels]{enumitem}
\usepackage{bm}
\usepackage{fullpage}
\usepackage{mathtools}

@@

% Absolute value
\DeclarePairedDelimiter\abs{\lvert}{\rvert}

\newcommand{\diam}{\mathrm{diam}}
\newcommand{\set}[1]{\left\{#1\right\}}
\newcommand{\defeq}{\overset{\mathrm{def}}{=}}
\newcommand{\vol}{\mathrm{vol}}
\newcommand{\cut}{\mathrm{cut}}
% \newcommand{\abs}[1]{\left \lvert #1 \right \rvert}
\newcommand{\N}{\mathbb{N}}
\newcommand{\Reals}{\mathbb{R}}
\newcommand{\Rd}{\Reals^d}
\newcommand{\norm}[1]{\left\lVert#1\right\rVert}
\newcommand{\1}{\mathbf{1}}
\newcommand{\Phibf}{\mathbf{\Phi}}
\newcommand{\Psibf}{\mathbf{\Psi}}
\newcommand{\dist}{\mathrm{dist}}

%%% Vectors
\newcommand{\pbf}{p}        % removed bold font
\newcommand{\qbf}{\mathbf{q}}
\newcommand{\ebf}[1]{\mathbf{e}_{#1}}
\newcommand{\pibf}{\bm{\pi}}

%%% Matrices
\newcommand{\Abf}{\mathbf{A}}
\newcommand{\Xbf}{X}             % removed bold font 
\newcommand{\Wbf}{\mathbf{W}}
\newcommand{\Lbf}{\mathbf{L}}
\newcommand{\Dbf}{\mathbf{D}}
\newcommand{\Ibf}[1]{\mathbf{I}_{#1}}

@@

%%% Sets
\newcommand{\Cset}{\mathcal{C}}
\newcommand{\Aset}{\mathcal{A}}
\newcommand{\Asig}{\Aset_{\sigma}}
\newcommand{\Csig}{\Cset_{\sigma}}
\newcommand{\Sset}{\mathcal{S}}

%%% Graph quantities
\newcommand{\Cest}{\widehat{C}}

@@

\begin{abstract}
\vskip 0.1 in % RESOLVED
Spectral clustering methods are a family of popular nonparametric clustering
tools.  Recent works have proposed and analyzed \emph{local} spectral methods,
which extract clusters using locally-biased random walks around a user-specified
seed node.  Several authors have shown that local methods, such as personalized
PageRank (PPR), have worst-case guarantees for certain graph-based measures of
cluster quality.  In contrast to existing works, we analyze PPR in a traditional
statistical learning setup, where we obtain samples from an unknown
distribution, and aim to identify connected regions of high-density (density
clusters).  We introduce two natural criteria for cluster quality, and derive
bounds for these criteria when evaluated on empirical analogues of density
clusters. Moreover, we prove that PPR, run on a neighborhood graph, extracts
sufficiently salient density clusters. Finally, we provide empirical support of our theory.
\end{abstract}

@@

Let $\Xbf = \{x_1, \ldots, x_n\}$ be a sample drawn i.i.d.\ from a
distribution $\Pbb$ on $\Rd$, with density $f$, and consider the problem of 
clustering: splitting the data into groups which satisfy some notion of
within-group similarity and between-group difference.  We focus on spectral
clustering methods, a family of powerful nonparametric clustering algorithms.
Roughly speaking, a spectral technique first constructs a geometric graph $G$,
where vertices are associated with samples, and edges correspond to proximities
between samples. It then learns a feature embedding based on the Laplacian of
$G$, and applies a simple clustering technique (such as k-means clustering) in
the embedded feature space.

To be more precise, let $G=(V,E,w)$ denote a weighted, undirected graph  
constructed from the samples $\Xbf$, where $V=\{1,\ldots,n\}$, and $w_{uv}
= K(x_u,x_v) \geq 0$ for $u,v \in V$, and a particular kernel function $K$.
Here $(u,v) \in E$ if and only if $w_{uv} > 0$.  We denote by $\Abf \in
\Reals^{n \times n}$ the weighted adjacency matrix, which has entries
$A_{uv}=w_{uv}$, and by $\Dbf$ the degree matrix, with 
\smash{$\Dbf_{uu} = \sum_{v \in V} \Abf_{uv}$}.  We also denote by $\Wbf,\Lbf$
the (lazy) random walk transition probability matrix and normalized\footnote{Other
  popular choices here include the unnormalized Laplacian, and symmetric
  normalized Laplacian.} 
Laplacian matrix, respectively, which are defined as
$$
\Wbf = \frac{\Ibf{} + \Dbf^{-1}\Abf}{2}, \quad \Lbf = \Ibf{} - \Wbf,
$$
where $\Ibf{} \in \Reals^{n\times n}$ is the identity matrix.  Classical global
spectral methods take a eigendecomposition $L=U \Sigma U^T$, use some 
number of eigenvectors (columns in $U$) as a feature representation for the
samples, and then run (say) k-means in this new feature space.

@@

For subsets $S,S' \subseteq V$, we define the cut, degree, and volume functionals as usual,
$$
\cut(S, S'; G) = \sum_{u \in S} \sum_{v \in S'} w_{uv}, \quad \deg(u; G) = \sum_{v \in V} w_{uv}, \quad  \vol(S; G) = \sum_{u \in S} \deg(u;G).
$$
As our notion of external connectivity, we use the \emph{normalized cut}  
of $S$, defined as 
\begin{equation}
\label{eqn: norm_cut}
\Phi(S; G) = \frac{\cut(S; G)}{\min\{\vol(S; G), \vol(S^c; G)\}},
\end{equation} 
where we abbreviate $\cut(S; G) = \cut(S; S^c; G)$.

@@

For $S \subseteq V$, denote by $G[S] = (S, E_S, w_S)$ the subgraph induced by 
$S$ (where the edges are $E_S = E \cap (S \times S)$). Let $\Abf_S,\Dbf_S$
be the adjacency matrix and degree matrix, respectively, of $G[S]$.  Define the
(lazy) random walk matrix as usual, $\Wbf = \frac{\Dbf_S^{-1} \Abf_S + \Ibf{\abs{S} \times \abs{S}}}{2}$, and for $v \in V$,
write 
$$
q_{v}^{(t)}(u) = e_v\Wbf_S^t e_u
$$
for the $t$-step transition probability of a random walk over $G[S]$
originating at $v$.\footnote{Given a starting node $v$ and and a random walk
  defined by transition probability matrix $\mathbf{P}$, the notation $e_v
  \mathbf{P}^t$ is used to denote the distribution of the random walk after $t$
  steps.}   Also write \smash{$\pi = (\pi(u))_{u \in S}$}
for the stationary distribution of this random walk.  (Given the definition of 
$\Wbf_S$, it is well-known that a unique stationary distribution exists and is given by
\smash{$\pi(u) = \deg(u; G[S])/\vol(S; G[S])$}.)

Our internal connectivity parameter will capture the time it takes for the
random walk over $G[S]$ to mix (approach the stationary distribution)
uniformly over $S$.  For this, we first define the \emph{relative pointwise mixing
  time} of $G[S]$ as 
$$
\tau_{\infty}(G[S]) = \min\set{ t: \frac{\pi(u) - q_{v}^{(t)}(u)
	}{\pi(u)} \leq \frac{1}{4}, 
	\; \text{for $u,v \in V$}}. 
$$
%where \smash{$\qbf = (\qbf_v^{(1)}, \qbf_v^{(2)}, ...)_{v \in V}$}, and
%\smash{$\qbf_v^{(m)} = (q_{vu}^{(m)})_{u \in V}$}. 
% RJT: I got rid of the notational dependence on \qbf, because these quantities
% are in turn just defined in terms of the induced subgraph G[S]
% AJG: RESOLVED
Now our internal connectivity parameter is simply the inverse mixing time, 
\begin{equation}
\label{eqn: inv_mixing_time}
\Psi(S; G) = \frac{1}{\tau_{\infty}(G[S])}.
\end{equation}

@@

We now describe the clustering algorithm that will be our focus for the rest of 
the paper. We start with the geometric graph that we form based on the samples 
$\Xbf$: for a radius $r > 0$, we consider the \emph{$r$-neighborhood graph} of 
$\Xbf$, denoted $G_{n,r}=(V,E)$, an unweighted graph with vertices
$V=\Xbf$, and an edge $(x_i,x_j) \in E$ if and only if $\norm{x_i - x_j}
\leq  r$, where $\norm{\cdot}$ denotes Euclidean norm.  Note that this is a
special case of the general construction introduced above, with 
$K(u,v) = 1(\norm{x_u - x_v} \leq r)$.

@@

For a level $\beta > 0$ and a target volume $\vol_0 > 0$, we define a
\emph{$\beta$-sweep cut} of $\pbf = (p_u)_{u \in V}$ as  
\begin{equation}
\label{eqn: sweep_cuts}
S_\beta = \{u \in V: \frac{p_u}{\Dbf_{uu}} > \frac{\beta}{\vol_{0}}\}.
\end{equation}
Having computed sweep cuts over a range \smash{$\beta \in  
(\frac{1}{40},\frac{1}{11})$},\footnote{The choice of a specific range such as 
\smash{$(\frac{1}{40}, \frac{1}{11})$} is standard in the analysis of PPR
algorithms, see, e.g., \citep{zhu2013}.}
we output a cluster \smash{$\Cest = S_{\beta^*}$}, based on the sweep cut
$S_{\beta^*}$ that minimizes the normalized cut \smash{$\Phi(S_{\beta^*};
  G_{n,r})$} as defined in \eqref{eqn: norm_cut}. For concreteness, we summarize
this procedure in Algorithm \ref{alg: ppr}.

\begin{algorithm}
\caption{PPR on a Neighborhood Graph}
\label{alg: ppr}	
{\bfseries Input:} data $\Xbf=\{x_1,\ldots,x_n\}$, radius $r > 0$, teleportation 
parameter $\alpha \in [0,1]$, seed $v \in \Xbf$, target stationary volume $\vol_0 >
0$. \\   
{\bfseries Output:} cluster $\Cest \subseteq V$.
\begin{algorithmic}[1]
  \STATE Form the neighborhood graph $G_{n,r}$.
  \STATE Compute the PPR vector $\pbf(v, \alpha; G_{n,r})$ as in \eqref{eqn:
    ppr_vector}. 
  \STATE For $\beta \in (\frac{1}{40}, \frac{1}{11})$ compute sweep cuts
  $S_{\beta}$ as in \eqref{eqn: sweep_cuts}.
  \STATE Return \smash{$\Cest = S_{\beta^*}$}, where 
  $$
  \beta^* = \argmin_{\beta \in (\frac{1}{40}, \frac{1}{11})} \Phi(S_{\beta}; G_{n,r}).
  $$
\end{algorithmic}
\end{algorithm}

@@

\begin{definition}[Misclassification error]
\label{def: misclassification_rate}
For an estimator \smash{$\Cest \subseteq \Xbf$} and set
$\mathcal{S} \subseteq \Reals^d$, the \emph{misclassification error} of $\mathcal{S}$ by $\Cest$ is
\begin{equation}
\label{eqn: misclassification_rate}
\abs{\Cest \setminus (\mathcal{S} \cap \Xbf)} + \abs{(\mathcal{S} \cap \Xbf) \setminus \Cest}.
\end{equation}
\end{definition}

\begin{definition}[Consistent density cluster estimation]
\label{def: consistent_density_cluster_estimation}
For an estimator \smash{$\Cest \subseteq \Xbf$} and cluster 
$\Cset \in \Cbb_f(\lambda)$, we say \smash{$\Cest$} is a consistent
estimator of $\Cset$ if for all $\Cset' \in \Cbb_f(\lambda)$ with $\Cset \not=
\Cset'$ the following holds as $n \to \infty$: 
\begin{equation}
\label{eqn: consistent_density_cluster_recovery}
\Cset[\Xbf] \subseteq \Cest \quad \text{and} \quad
\Cest \cap \Cset'[\Xbf] = \emptyset,
\end{equation}
with probability tending to 1.
\end{definition}

@@

\item Under additional geometric conditions, which exclude sets with large diameter or small bottleneck, we derive in Theorem
  \ref{thm: mixing_time_upper_bound} a lower bound on the
  inverse mixing time of a random walk over $\Csig[\Xbf]$.

\item In Section \ref{sec: consistent_cluster_estimation_with_ppr}, we show in
  Theorems \ref{thm: misclassification_rate} and \ref{thm: consistent_recovery_of_density_clusters} that the bounds on the cluster quality criteria established in Theorems \ref{thm: conductance_upper_bound} and \ref{thm:
    mixing_time_upper_bound} have algorithmic consequences for \ppr. 
  Properly initialized, Algorithm \ref{alg: ppr} has low misclassification error with respect to a small enlargement of the set $\Cset$, and if the density cluster $\Cset$ is particularly well-conditioned, Algorithm \ref{alg: ppr} will perform consistent density cluster estimation in the sense of \eqref{eqn: consistent_density_cluster_recovery}. Corollary \ref{cor: appr} establishes that these statements hold also with respect to an approximate form of \ppr, which can be efficiently computed.

\item In Section \ref{sec: experiments}, we empirically
  demonstrate the tightness of the bounds in Theorems \ref{thm: conductance_upper_bound} and \ref{thm: mixing_time_upper_bound}, and provide examples showing how violations of the geometric conditions we require manifestly
  impact density cluster recovery by \ppr.  
\end{enumerate}

@@

In addition to the background given above, a few related lines of work are worth
highlighting.
Global spectral clustering methods were first developed in the
context of graph partitioning \citep{fiedler1973,donath1973} and their
performance is well-understood in this context (see, e.g.,
\citep{tolliver2006,luxburg2007}).  In a similar vein, several recent works
\citep{mcsherry2001,rohe2011,kamalika2012,balakrishnan2011,lei2015,abbe2018} 
have studied the efficacy of spectral methods in successfully recovering the
community structure in the stochastic block model and variants.

@@

\item
\label{asmp: embedding}
\emph{Lipschitz embedding:}
There exists $\mathcal{K} \subseteq \Rd$ convex, and $g: \Reals^d \to \Reals^d$ satisfying, for some $ L \geq 1$,
\begin{equation*}
\det(\nabla g (x) ) = 1,\frac{1}{L}\norm{x - y} \leq \norm{g(x) - g(y)} \leq L \norm{x - y} ~ \text{for all $x,y \in \Reals^d$}
\end{equation*}
such that $\Csig$ is the image of $\mathcal{K}$ by $g$, $\Csig = g(\mathcal{K})$.
Furthermore, there exists $D < \infty$ such that for all $x, x' \in \mathcal{K}$
$$
\norm{x - x'} \leq D.
$$
\end{enumerate}

Note that $\sigma$ plays several roles here, precluding arbitrarily narrow
clusters in \ref{asmp: bounded_density}, flat densities around the level set in \ref{asmp: low_noise_density}, and poorly separated clusters in \ref{asmp: cluster_separation}.

Assumptions \ref{asmp: bounded_density}, \ref{asmp: low_noise_density},
and \ref{asmp: cluster_separation} are used to upper bound $\Phi(\Cset[\Xbf];
G_{n,r})$, whereas \ref{asmp: bounded_density}, and \ref{asmp: embedding} are required to lower bound $\Psi(\Cset[\Xbf]; G_{n,r})$. We note that the
lower bound on minimum density in \ref{asmp: bounded_density} along with \ref{asmp:  
cluster_separation} are similar to the $(\sigma,\epsilon)$-saliency of
\citep{chaudhuri2010}, a standard density clustering assumption, while
\ref{asmp: low_noise_density} is seen in, e.g., \citep{singh2009} (as well as
many other works on density clustering and level set estimation.) While \ref{asmp: embedding} may be less standard, as we will see, it is critical in order to achieve reasonably tight bounds on $\Psi(\Cset[\Xbf];
G_{n,r})$.  It is also worth
highlighting that these assumptions are all local in nature, a benefit of
studying a local algorithm such as \ppr.

@@

We emphasize that while many of these geometric conditions are typical in the density clustering literature, the restrictions we will impose upon them in order to obtain meaningful implications for \pprspace will not be. This is natural. The spectral algorithm we consider is not specifically designed for the task of level set estimation, and in fact one should expect \pprspace to fail to recover -- either in the sense of \eqref{eqn: consistent_density_cluster_recovery}, or indeed any reasonable notion of cluster recovery -- a density cluster of sufficiently large diameter or sufficiently small thickness (though we do not provide any lower bounds to this effect). Indeed, one of the primary motivations of this work was to better understand and characterize the distinctions between those level sets which are well conditioned for spectral algorithms, and those which are not.

@@

\begin{remark}
The inequality in \eqref{eqn: conductance_additive_error_bound} is tight in the case of $\Cset = \set{0}$. To see this, let $\Csig = B(0,\sigma)$ and
$$
 f(x)  = \begin{cases}
   \lambda &\text{for $x \in \Csig$}, \\
   \lambda - \dist(x,\Csig)^{\gamma} & \text{ for $0 < \dist(x,\Csig) < r$}, 
\end{cases}
$$
Then, some simple calculations yield
\begin{equation*}
\Ebb(\cut(\Csig[\Xbf]; G_{n,r})) \geq c \lambda \nu_d r^d \Pbb((\Csig + B(0,r)), \quad \textrm{and} \quad \Ebb(\vol_{n,r}(\Csig[\Xbf]; G_{n,r}) \leq c' \lambda \nu_d r^d \Pbb(\Csig)
\end{equation*}
for some constants $c,c' > 0$. Thus the ratio $\Ebb(\cut(\Csig[\Xbf]; G_{n,r})) / \Ebb(\vol_{n,r}(\Csig[\Xbf]; G_{n,r})$ matches \eqref{eqn: conductance_additive_error_bound}, up to constants.
